---
title: "Multiple Regression"
author: "Modern Data Mining"
output:
  html_document:
    code_folding: show
    highlight: haddock
    number_sections: yes
    theme: lumen
    toc: yes
    toc_depth: 4
    toc_float: yes
  pdf_document:
    extra_dependencies: ["dcolumn"]
    number_sections: yes
    toc: yes
    toc_depth: '4'
  word_document:
    toc: yes
    toc_depth: '4'
urlcolor: blue
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, results = "hide", fig.width=8, fig.height=4)
# get the output format for stargazer only:
# rsults = `asis`
# stargazer( type = ) one of the following:
## if run in studio, set as "text"
## if knit to PDF, set as "latex"
## if knit to HTML, set as "html"
# We made it automatic:
output_format <- ifelse(is.null(knitr::opts_knit$get("rmarkdown.pandoc.to")),
                        "text", knitr::opts_knit$get("rmarkdown.pandoc.to"))

options(scipen = 0, digits = 3)  # controls base R output
if(!require('pacman')) {
  install.packages('pacman')
}
pacman::p_load(dplyr, ggplot2, gridExtra, ggrepel, leaps, car, tidyverse, contrast, sandwich, stargazer)
# car::Anova(): report F statistics
# leaps::regsubsets(): model selection
# tidyverse is a package of packages: ggplot2, dplyr, readr...
# sandwich::coeftest(fm, df = Inf, vcov = vcovHC)
# output_format <- 'text'
```

\pagebreak


# Objectives {-}

Data Science is a field of science. We try to extract useful information from data. In order to use the data efficiently and correctly we must understand the data first. According to the goal of the study, combining the domain knowledge, we then design the study. In this lecture we first go through some basic explore data analysis to understand the nature of the data, some plausible relationship among the variables. 



Data mining tools have been expanded dramatically in the past 20 years. Linear model as a building block for data science is simple and powerful. We introduce/review linear models. The focus is to understand what is it we are modeling; how to apply the data to get the information; to understand the intrinsic variability that statistics have. 


Suggested reading:

* This lecture in details
* Chapter 3 of ISLR
* Stat Sleuth: Ch 7- 10 Multiple Reg, Ch 5: One Way ANOVA
* Note: Many important topics are in Appendices. Please go through them.
* Data needed: `car_04_regular.csv` and `Cars_04.csv`


1. Elements to multiple regression
      - Model specification
      - General linear models
      - LS estimates and their properties
      - Inference
          *for the coefficients
          *for the mean response
          *for a future response
      
2. Categorical predictors
      - Adding categorical predictors to the multiple regression (ANOVA)
      - Model with both continuous and categorical variables (ANCOVA)
      - Models with or without interactions
      
3. More about multiple Regression
      - Sandwich standard error estimates to handle heteroscedasticity vs. homoscedasticity
      - Modeling non-linear relationships
      - Transformation of variables to meet linear model assumptions 
      - Outliers, Leverage points
      
4. Summary

5. Appendices

6. R functions 
    + lm()
    + anova()
    + Anova()
    + sandwich, lmtest::coeftest: sandwich cov estimators 
    + stargazer (organize table, lm output)
    
\pagebreak

# Case Study: Fuel Efficiency in Automobiles

When an automobile manufacture designs a new car, what is the benchmark to compare with in terms of fuel efficiency? How well can we predict housing price based on the house characteristics? Multiple regression is an extension of simple regression. It explores relationship between one variable (response) with many other variables (predictors or features). We highlight the regression methods in this lecture. It is important to understand the effect of each predictor on the response depending on what else are included in the model. Our approach here unifies usual multiple regression and ANOVA. 


Goal of the study: How to build a fuel efficient car?

- Effects of features on a car
- Given a set of features, we'd like to estimate the mean fuel efficiency as well as the efficiency of one car
- Are Asian cars more efficient than cars built in other regions?

We will use the `car_04_regular.csv` data set to perform a case study on fuel efficiency of automobiles. Below is a quick description of all the features in our dataset.

<center>
|Feature|Description|
|----------|----------------------------------------------|
|Continent|Continent the Car Company is From|
|Horsepower|Horsepower of the Vehicle|
|Weight|Weight of the vehicle (thousand lb)| 
|Length|Length of the Vehicle (inches) |
|Width|Width of the Vehicle (inches) |
|Seating|Number of Seats in the Vehicle|
|Cylinders|Number of Engine Cylinders|
|Displacement|Volume displaced by the cylinders, defined as $\pi/4 \times bore^{2} \times stroke \times \text{Number of Cylinders}$|
|Transmission|Type of Transmission (manual, automatic, continuous)|
</center>

Once we know the data structure we will rephrase our goal of study.  Specifically, fuel efficiency is measured by Mileage per Gallon, $Y$ = `MPG_City` 

  1. Predictors: Effects of each feature on $Y$
  2. Estimate 
    - the mean `MPG_City` for all such cars specified below and
    - predict $Y$ for the particular car described below
  3. Are cars built by Asian more efficient?
  
  4. In particularly, we would like to investigate the `MPG_City` for this newly designed American car

<center>
|Feature|Value|
|----------|----------------------------------------------|
|Continent|America|
|Horsepower|225|
|Weight|4|
|Length|180|
|Width|80|
|Seating|5|
|Cylinders|4|
|Displacement|3.5|
|Transmission|automatic|
</center>

## Exploratory Data Analysis

It is always crucial to look at the data first. For the purpose of shortening the lecture time, we put this part in Appendix 1 which explains how we cleaned the original data `Car_04.csv` and created a new version called `car_04_regular.csv`.

To summarize the EDA quickly, we excluded some high-performance cars with large Horsepower `(HP > 400)`. We also take a subset of features, containing only the features described in the description table.

## A quick glimpse

```{r results=TRUE}
data1 <- read.csv("data/car_04_regular.csv", header=TRUE)

names(data1)
dim(data1)   # 226 cars and 13 variables
str(data1)
head(data1) #View(data1)
```

We create a new car for prediction later as follows.

```{r results=TRUE}
newcar <-  data1[1, ]  # Create a new row with same structure as in data1
newcar[1] <- "NA" # Assign features for the new car
newcar[2] <- "Am"
newcar["MPG_City"] <- "NA"
newcar["MPG_Hwy"] <- "NA"
newcar[5:11] <- c(225, 4, 180, 80, 5, 4, 3.5)
newcar$Make <-  "NA"
newcar[13] <- "automatic"
newcar
```


\pagebreak

# Introduction to Multiple regression


## Model Specification

How does `Length` affect `MPG_City`? 

It depends on how we model the response. We will investigate **three models** with `Length`.

For the ease of presentation, we define some predictors below that we will use in subsequent models:

$$x_1 = Length, \quad x_2 = Horsepower, \quad x_3 = Width, \quad x_4 = Seating, \quad x_5 = Cylinders, \quad x_6 =  Displacement$$

We will create three models:

**M1. Our first model will only contain one predictor, Length:**


$$y_i = \beta_0 + \beta_1 \cdot x_{i1} + \epsilon$$ 


**Interpretation of $\beta_1$** is that in general, the mean $y$ will change by $\beta_1$ if a car is 1" longer. So we can't really peel off the effect of the `Length` over $y$. 


**M2. Next, we add the predictor `Horsepower` to our model**
  
$$y_i = \beta_0 + \beta_1 \cdot x_{i1} + \beta_2 \cdot x_{i2} + \epsilon$$

**Interpretation of $\beta_1$** is that in general, the mean $y$ will change by $\beta_1$ if a car is 1" longer and the 'Horse Power's` are the same. 



**M3. Finally, we fit a model with multiple predictors**

$$y_i = \beta_0 + \beta_1 x_{i1} + \beta_2x_{i2} + \beta_3x_{i3} + \beta_4x_{i4} + \beta_5x_{i5} + \beta_6x_{i6} + \epsilon$$

**Interpretation of $\beta_1$** is that in general, the mean $y$ will change by $\beta_1$ if a car is 1" longer and the rest of the features are the same.  



**Question**: Are all the $\beta_1$s same in the 3 models above? 

**No**. The effect of `Length` $\beta_1$ depends on the rest of the features in the model!!!! 



## General multiple linear models

In general, We define a multiple regression as

$$Y = \beta_0 + \beta_1 x_{i1} + \beta_2 x_{i2} + \dots + \beta_p x_{ip} + \epsilon_i$$

* Linearity Assumption for this model is 

$$\textbf{E}(y_i | x_i) = \beta_0 + \beta_1 x_{i1} + \beta_2 x_{i2} + \dots + \beta_p x_{ip} $$

* The homoscedasticity assumption is

$$\textbf{Var}(y_i | x_{i1}, x_{i2}, \dots, x_{ip}) = \sigma^2$$

* The Error term is iid. and defined as

$$\epsilon_i \overset{iid}{\sim} \mathcal{N}(0, \sigma^2)$$
This is same to say the errors are independent and with a normal distribution of mean $0$ and an equal variance $\sigma^2$.


* Parameters of interests are all the $\beta's$ and $\sigma^2$.

- Simple interpretations of each $\beta_i's$
- Simple models for predictions



## OLS and its Properties

These $\beta$ parameters are estimated using the same approach as simple regression, specifically by minimizing the sum of squared residuals ($RSS$): 

$$\min_{b_0,\,b_1,\,b_{2},\dots,b_{p}} \sum_{i=1}^{n} (y_i - b_0 - b_1 x_{i1} - b_2 x_{i2} - \dots - b_p x_{ip})^{2}$$

To be specific: 

We define let $X$ denote a $n × (p+1)$ matrix whose $(i,j)$th element is $x_{ij}$. That is:

$$X=
 \begin{pmatrix}
 1&  x_{11} & x_{12} & \dots & x_{1p} \\
 1&  x_{21} & x_{22} & \dots & x_{1p} \\
\vdots & \vdots & \vdots & \ddots & \vdots \\
 1&   x_{n1} & x_{n2} & \dots & x_{np} 
 \end{pmatrix}$$ 
 

### OLS Estimates

**Mathematical formula for the OLS estimates:**

The least squared estimators can be obtained precisely by  $$\hat\beta = (X^{T}X)^{-1} X^{T}Y $$

**Variance of the OLS:**

The variance-covariance matrix of $\hat \beta$

$$\textbf{V}[\hat\beta | X] = \sigma^{2} (X^{T}X)^{-1}$$
**Remark:**

The above covariance matrix is obtained with a homeostatic variance of $y's$. 

**Normality of the OLS:**

Under all the linear model assumptions, the LSE's are unbiased estimators of $\beta$ and they are normally distributed. 
$$ \hat\beta | X \sim N(0, \sigma^{2} (X^{T}X)^{-1}$$)

**Standard errors**

As soon as we know how to estimate common $\sigma^2$, we can estimate $\textbf{V}[\hat\beta | X]$ by 
$\hat {\textbf{V}}[\hat\beta | X] = \hat \sigma^{2} (X^{T}X)^{-1}$. The standard error for each $\hat \beta_i$ will be
$$\text {se}\, (\hat \beta_i) = \sqrt{( \hat \sigma^{2} (X^{T}X)^{-1} )}|_{ith\,\text{Diag}}$$

**Confidence intervals each $\beta_i$:**

- a $95\%% t-interval for $\beta_i$ is 
$$\hat \beta_i \pm t^*_{.025} \text{se}(\hat \beta_i).$$
- Here $t^*_{.025}$ is the upper .025 percentile from a $t$ distribution. You may take it to be approximately $1.96$ if we have have a large number of observations $n$. 

- $\text{se}(\hat \beta_i)$ is taken from the variance formula above.

- It is a t-confidence interval because we need to estimate the unknown $\sigma^2$.

**Hypothesis test for each $\beta_i$:**

To test that 
$$\beta_i = 0  \;\;\; \text{vs.}\;\;\; \beta_i = 0$$ 

which means that given other variables in the model, there is no $x_i$ effect. We carry out a t-test:

$$ \text{tstat} = \frac{\hat \beta_i - 0}{\text{se} (\hat \beta_i)}$$
The p-value is: 
$$\text{p-value} = 2 \times P(\text{T variable} > \text{tstat})$$.

We reject the null hypothesis at an $\alpha$ level if the p-value is < $\alpha$. 

**A $95\%$ Confidence interval for the mean given a set of predictors:**

$$\hat y \pm \,\,t^*_{.025}\,\, \times se(\hat y) $$


**A $95\%$ prediction interval for a future $y$ given a set of predictors:**
$$\hat y \pm \,\,t^*_{.025}\,\, \times \hat \sigma.$$


### Residual Sum of Squares (RSS)
 
For multiple regression, $RSS$ is estimated as:

$$RSS =  \sum_{i=1}^{n} \hat{\epsilon}_i^{2} = \sum_{i=1}^{n} (y_i-\hat y_i)^2= \sum_{i=1}^{n} (y_i - (\hat\beta_0 + \hat\beta_1 x_{i1} + \hat\beta_2 x_{i2} + \dots + \hat\beta_p x_{ip}))^{2}$$

### Mean Sum of Squares (MSE), Residual Standard Error (RSE)

$\sigma^2$ is estimated by the Mean Sum of Squares (MSE). For multiple regression, MSE is defined as:


$$MSE = \frac{RSS}{n-p-1} \\ 
\text{where }p =\text{number of predictors}$$


$\sigma$ is estimated by the Residual Standard Error (RSE). For multiple regression, RSE is defined as:

$$\hat \sigma= RSE = \sqrt{MSE} = \sqrt{\frac{RSS}{n-p-1}}$$
**Remark:**

Notice the denominator in $MSE$ is $n-(p+1)$, where $p+1$ is the number of predictors in the model +1. By doing so we have a nice mathematical property: $RSE$ is an unbiased estimator of $\sigma^2$. In another word, in a long run, on average the estimator hits the quantity being estimated. 


### Goodness of Fit: $R^2$

**Total Sum of Squares (TSS)**: 

TSS measures the total variance in the response $Y$. When all the responses are the same which means no predictors affect the response values, then $TSS=0$. In contrast, RSS measures the amount of variability that is left unexplained after performing the regression.

$$TSS = \sum_{i=1}^{n} (y_i - \bar{y})^{2}$$


**$R^2$**: 

How much variability is captured in the linear model using this set of predictors? $R^{2}$ measures the proportion of variability in $Y$ that can be explained using this set of predictors in the model. 

$$R^{2} =  \frac{TSS - RSS}{TSS}$$
**Remark 1:**

  *$TSS \geq RSS$. Why so???
  *$R^2 \leq 1$. 
  *$TSS=RSS + \sum (\hat y_i - \bar y) ^2$.
  *$(corr (y, \hat y))^2$

An $R^{2}$ statistic that is close to 1 indicates that a large proportion of the variability in the response has been explained by the regression.


**Remark2**:

*How large $R^2$ needs to be so that you are comfortable to use the linear model?
*Though $R^2$ is a very popular notion of goodness of fit, but it has its limitation. Mainly all the sum of squared errors defined so far are termed as `Training Errors`. It really only measures how good a model fits the data that we use to build the model. It may not generate well to unseen data. 



## R function `lm()

Linear models are so popular due to their nice interpretations as well as clean solutions.  R-function `lm()` will be. It takes a model specification together with other options, outputs all the estimators, summary statistics such as varies sum of squares, standard errors of estimators, testing statistics and  p-values.  Finally the predicted values with margin of errors or confidence intervals and prediction intervals can be called.  


## Compare three models
Linear model effects or coefficients are defined within the model, they depend on the rest of the variables in the model.  Let us compare the three model fits, focusing on the interpretations over the coefficients of `length` in different models. 

### Model 1: `MPG_City ~ Length

We first do a simple regression. Let the response $y_i$ be the `MPG_City` and the explanatory variable $x_{i1}$ be `Length` ($i = 1, \dots, n=226$).


$$y_i = \beta_0 + \beta_1 \cdot x_{i1} + \epsilon$$

Model assumptions:

1. Linearity: given `Length`, the mean of `MPG_City` is described as $\textbf{E}(y_i | x_i) = \beta_0 + \beta_1 x_i$
2. Equal variance: the variances of `MPG_City` are the same for any `Length` i.e. $\textbf{Var}(y_i | x_{i1}) = \sigma_{1}^2$
3. Normality: `MPG_City` is independent and normally distributed. i.e. $\epsilon_i \overset{iid}{\sim} \mathcal{N}(0, \sigma_{1}^2)$


We now create a model with `lm()`
```{r}
fit1 <- lm(MPG_City ~ Length, data = data1)    # model specification response ~ x1,.. 
ggplot(data1, aes(x = Length , y = MPG_City)) + 
  geom_point() +
  geom_smooth(method="lm",se=F) + 
geom_hline(aes(yintercept = mean(MPG_City)), color = "red") 
```

Note from the summary below, the $\hat\beta$ for Length is estimated as `r fit1$coefficients[2]`. We say on average MPG drops $.13983$ if a car is $1$'' longer.
```{r results=TRUE}
fit1 <- lm(MPG_City ~ Length, data = data1)  # model one 
summary(fit1) 
# predict(fit1, newcar, interval ="prediction",  se.fit = TRUE)  
# TSS <- sum((data1$MPG_City - mean(data1$MPG_City))^2)
# RSS <- sum((fit1$res)^2); RSS
# model diagnosese
# plot(fit1$fit, fit1$res); same as plot(fit1, 1) # linearity and homoscedasticity
# qqnorm(fit1$res); qqline(fit1$res) # looking for a well fitted straight line
# plot(fit1, 2)
```


### Model 2: `MPG_City ~ Length + Horsepower`

Let the response $y_i$ be the `MPG_City` and the explanatory variables be `Length` and `Horsepower` ($i = 1, \dots, n=226$).


$$y_i = \beta_0 + \beta_1 \cdot x_{i1} + \beta_2 \cdot x_{i2} + \epsilon$$


```{r, results='asis'}
fit2 <- lm(MPG_City ~ Length + Horsepower, data = data1) 
stargazer::stargazer(fit2, type=output_format, align=TRUE)
```


A couple quick notes:

1. The order for predictors plays no role.
2. The output of `lm()` is similar regardless how many predictors are used.
3. The model assumptions extended automatically here. 

```{r results=TRUE}
summary(fit2)  #sum((fit2$res)^2)
```



**NOTICE**: Comparing `fit2: Length + Horsepower` to `fit1: Length` 

1. The coefficient for `Length` changed to $-0.061651$ from $-0.13983$!!!!
2. $R^{2}$ is $0.5908$ from $0.266$ - a huge increase. (Never decreasing, why?)
3. $RSE$ (Residual standard error) is now $2.372$ decreased from $3.175$ (almost never decreasing, why?)


The effects of variables ($\beta's$) are defined within the model. They all depend on what else are included or accounted for!


### Model 3:  Several continuous variables

We now add multiple variables to our model:

$$y_i = \beta_0 + \beta_1 x_{i1} + \beta_2x_{i2} + \beta_3x_{i3} + \beta_4x_{i4} + \beta_5x_{i5} + \beta_6x_{i6} + \epsilon$$


```{r results=TRUE}
fit3 <- lm(MPG_City ~ Length + Horsepower + Width + Seating +
           Cylinders + Displacement, data = data1)
summary(fit3)  
# shift+c+command: to comment the portion selected 
# confint(fit3)
# 2*pnorm(2.84, lower.tail = FALSE) # pvalues
# newcar
# predict(fit3, newcar)
# RSS <- sum((fit3$residual)^2)
# MSE <- RSS/219
# RSE <- sqrt(MSE)   
```


Based on the output from model 3:

 $\hat\beta_{Length}$ = `r coef(fit3)[2]` for model 3. Is this estimate wrong?



To summarize, the effects of `Length` from the above three models are

```{r results='asis'}
data.frame(Model.1 = coef(fit1)[2], Model.2 = coef(fit2)[2], Model.3 = coef(fit3)[2])
# process the lm output into a table
stargazer(fit1, fit2, fit3, type = output_format, 
                     # ci=TRUE, ci.level = .95,
                     keep.stat = c("n", "rsq", "sigma2", "ser"))
```

* They are different as expected
* Each one has its own meaning!



## Inference for a fixed model


Given a fixed model with several predictors, say model 3, assume all the linear model assumptions are met. Let us walk through the possible outcomes and point out the limitations. 


**Fit the model:**


```{r results=TRUE}
fit3 <- lm(MPG_City ~ Length + Horsepower + Width + Seating +
           Cylinders + Displacement, data = data1)
summary(fit3) ### key output
# shift+c+command: to comment the portion selected 
# confint(fit3)
# 2*pnorm(2.84, lower.tail = FALSE) # pvalues
# newcar
# predict(fit3, newcar)
# RSS <- sum((fit3$residual)^2)
# MSE <- RSS/219
# RSE <- sqrt(MSE)
```

```{r}
## clean summary table and arrange it by tstat
fit3.summary <- summary(fit3)   #names(summary(fit3))
    data.frame(fit3.summary %>% coefficients) %>%
    rename(estimate = Estimate,
          se = "Std..Error",
          tstat = "t.value",
          pvalue = "Pr...t..") %>%
    mutate(abs.tstat = abs(tstat)) %>%
  arrange(desc(abs.tstat)) %>%
knitr::kable()
```


**Questions of interests:**

Students please fill in the answers with output from R. 

1. Write down the final OLS equation of `MPG_City` given the rest of the predictors.

2. What is the standard error from the output? Precisely what does it measure?

3. Interpret the $R^2$ reported for this model. Do you feel comfortable using the output for the following questions based on this $R^2$ value?

4. What does each t-interval and t-test do?

5. Can you conclude that with $95\%$ confidence all the t confidence intervals provided above are right simultaneously?

6. Is `Width` **THE** most important variable, `HP` the second, etc since they each has the smallest p-value,...

7. Can we conclude that none of the `Seating` or `Cylinders` is needed?

8. Is `Width` most useful variable due to its largest coefficient in magnitude? 

9. Can we conclude that none of the `Seating` or `Cylinders` is needed?




## Confidence Interval for the Mean

Base on Model 3, the mean of `MPG_City` among all cars with the same features as the new design: length=180, HP=225, width=80, seating=5, cylinders=4, displacement=3.5, transmission="automatic", continent="Am" is

$$\hat{y} = 45.63 + 0.05 \times 180 - 0.02 \times 225 - 0.35 \times 80 - 0.24 \times 5 - 0.27 \times 4 - 0.94 \times 3.5 = 16.17, $$
with a $\text{standard error}$ for the regression line = 0.7839001.

We will get the prediction, standard error and the confident intervals by `predict()`.

```{r results=TRUE}
predict(fit3, newcar, interval = "confidence", se.fit = TRUE) 
```

Q: What assumptions are needed to make this a valid confidence interval?


## Prediction Interval 

Base on Model 3, `MPG_City` for this particular new design is

$$\hat{y} = 45.63 + 0.05 \times 180 - 0.02 \times 225 - 0.35 \times 80 - 0.24 \times 5 - 0.27 \times 4 - 0.94 \times 3.5 = 16.17$$
with a 95% prediction interval approximately to be 

$$\hat{y} \pm 2\times RSE = 16.17 \pm 2 \times 2.036.$$

The prediction interval can be obtained by `predict()`.

```{r results=TRUE}
predict(fit3, newcar,  interval = "predict", se.fit = TRUE) # future prediction intervals
```

Q: What assumptions are needed to make this a valid prediction interval?


## Model Diagnoses
To check the model assumptions are met, we examine the residual plot and the qqplot of the residuals.

We use the first and second plots of `plot(fit)`.

```{r eval=T}
par(mfrow=c(1,2), mar=c(5,2,4,2), mgp=c(3,0.5,0)) # plot(fit3) produces several plots
plot(fit3, 1, pch=16) # residual plot
abline(h=0, col="blue", lwd=2)
plot(fit3, 2) # qqplot
```
Are the linear model assumptions met for the model fit here (fit3)? What might be violated? 
-linearity?
-Equal variances?
-Normality?


# F-test

With multiple variables, we need to test whether a set of $\beta$'s being 0. 
For example, we want to ask whether $\beta_\text{Seating}$ and $\beta_\text{Cylinder}$ are not useful.
Mathematically, we test the null hypothesis

$$H_0: \beta_\text{Seating}=\beta_\text{Cylinder}=0 \mbox{ vs. } H_1: \text{at least one of } \beta_\text{Seating},\beta_\text{Cylinder} \text{ is non-zero}$$
    
To put it in another way, $H_0$ is the **reduced model:**

$$y_i = \beta_0 + \beta_{\text{Length}} x_{\text{i,Length}} + \beta_{\text{HP}}x_{\text{i,HP}} + \beta_{\text{Width}}x_{\text{i, Width}} + \beta_{\text{Displacement}}x_{\text{i, Displacement}} + \epsilon_i$$

and $H_1$ is the **full model:**

$$y_i = \beta_0 + \beta_{\text{Length}} x_{\text{i,Length}} + \beta_{\text{HP}}x_{\text{i,HP}} + \beta_{\text{Width}}x_{\text{i, Width}} + \\
 \beta_{\text{Seating}}x_{\text{i, Seating}} + \beta_{\text{Cylinders}}x_{\text{i, Cylinders}} + \beta_{\text{Displacement}}x_{\text{i, Displacement}} + \epsilon_i$$


**Theorem**: Assume the linear model assumptions hold, then under the $H_0$

$$F_{\text{stat}}:=\frac{\frac{RSS(H_0)-RSS(H_1)}{df_1}}{\frac{RSS(H_1)}{df_2}} \sim F_{df_1, df_2}$$
where 
$$\begin{split} 
{df}_1 &= \text{\# of parameters in } H_1 - \text{\# of parameters in } H_0 \\
{df}_2 &= \text{\# of observations} - \text{\# of parameters in } H_1
\end{split}.$$

The p-value is

$$\text{p-value}=P(F_{{df}_1, {df}_2} > F_{\text{stat}})$$

To explore, we can run a model without `Seating` and `Cylinder`.
```{r}
fit3.0 <- lm(MPG_City ~ Horsepower + Length + Width + Displacement, data = data1) # under H0
sum((summary(fit3.0)$residuals)^2) # RSS(H_0)
sum((summary(fit3)$residuals)^2) # RSS(H_1)
```

$$F_{\text{stat}}=\frac{\frac{RSS(H_0)-RSS(H_1)}{df_1}}{\frac{RSS(H_1)}{df_2}} = \frac{\frac{923.82 - 907.76}{2}}{\frac{907.76}{219}}=1.9373.$$

Based on the $F$-test given above, the $p$-value is 
$$P(F_{2,219} > 1.9373)$$

```{r results=TRUE}
pf(1.9373, 2, 219, lower.tail=F)  
# hist(rf(10000, 2, 219), breaks=40)    # take a look at the F dis
```

## `anova()`
We can use the `anava()` function to carry out the above F-test `anova(H_0, H_1)`.

```{r results=TRUE}
fit3.0 <- lm(MPG_City ~ Horsepower + Length + Width + Displacement, data = data1)
anova(fit3.0, fit3)
```

**Conclusion:** The p-value being .1466 gives us no evidence to reject the null hypothesis.



Questions based on the summary for `fit3`

1) What is the $F$ statistics from `summary(fit3)`?

2) Can you perform a $F$-test for each $\beta$ being 0? What is the relationship between the F-test and the t-test?

```{r R^2 and F}
TSS <- sum((data1$MPG_City - mean(data1$MPG_City))^2)   # (226-1)*var(data1$MPG_City)
TSS


```


Roughly speaking, the linearity, homoscedasticity and normality assumptions are satisfied. (Very subjective!)


\pagebreak

# Categorical Predictors

Let's use `Continent` as one variable. It has three categories. We explore the following questions:

1. Are Asian cars more efficient? 
2. How does `Continent` affect the `MPG`?

```{r results=TRUE}
levels(data1$Continent)   #data1$Continent
```

Model with a categorical variable is termed as a "One Way ANOVA"

First, we get the sample means and sample standard error for each group, and plot each
```{r results=TRUE}
data1 %>%
group_by(Continent) %>%
  summarise(
    mean = mean(MPG_City),
    sd  = sd(MPG_City),
    n = n()
  )
  
ggplot(data1) + geom_boxplot(aes(x = Continent, y = MPG_City))
```


We use indicator predictors to analyze the effect of `Continent`.

## Model 1 - One Way ANOVA

$$y_{i|x=Am} = \mu_{Am} + \epsilon_i$$
$$y_{i|x=As} = \mu_{As} + \epsilon_i$$
$$y_{i|x=E} = \mu_{E} + \epsilon_i$$
where $\mu_{Am}$, $\mu_{As}$ and $\mu_{E}$ are the mean MPG of each continent and $\epsilon \sim \mathcal{N}(0,\sigma^2)$. We want to compare the three means and this can be done through linear model with indicator functions.

We are interested in the following hypotheses:

$$H_0: \mu_{Am} = \mu_{As} = \mu_{E}$$


Let 

* `Am` as the base 

* $x_1$ be the indicator function of being `As`, i.e. $I\{Continent = As\}$

* $x_2$ be the indicator function of being `E` $I\{Continent = E\}$

Then the above Anova model is same as

$$y_i = \beta_{Am} + \beta_{As} x_{1,i} + \beta_E x_{2,i} + \epsilon_i = \beta_{Am} + \beta_{As} I\{Continent = As\} + \beta_E I\{Continent = E\} + \epsilon_i$$
where
* $\beta_{Am} = \textbf{E}(MPG|Am)$

* $\beta_{As} = \textbf{E}(MPG|As)-\textbf{E}(MPG|Am)$: the increment between Asia and America

* $\beta_E = \textbf{E}(MPG|E)-\textbf{E}(MPG|Am)$: the increment between Europe and America


To test $$H_0: \mu_{Am} = \mu_{As} = \mu_{E}$$ is same as to test
$$H_0: \beta_{As} = \beta_{E} = 0$$


Now we can use `lm()` to fit the model and carry out the tests

```{r results=TRUE}
fit.continent <- lm(MPG_City ~ Continent, data1)
summary(fit.continent)  #confint(fit.continent); Anova(fit.continent); anova(fit.continent)
```

The $F_{test}$ here is to test $H_0: \; \beta_{As} = \beta_{E} = 0$, i.e., there is no difference among the three regions. We reject $H_0$ at $\alpha=0.01$, and conclude the mean MPG's are different among the three regions.

We can also see the coding from here that Am is the base, each coefficient captures the effect of the level vs. Am.
```{r}
model.matrix(fit.continent)[5:10, ]
```

Remarks: `fit.continent` can also be obtained from
```{r}
fit.continent.1 <- lm(MPG_City ~ I(Continent == "Am") + I(Continent == "E") , data1)
summary(fit.continent.1)    
model.matrix(fit.continent.1)[5:10, ]
```

We also show a model without intercept $\beta_0$. Can you interpret the summary table below? 
```{r results=TRUE}
fit.continent.2 <- lm(MPG_City ~ 0 + Continent , data1) #
summary(fit.continent.2)  
model.matrix(fit.continent.2)[5:10, ] # try to show you the indicator variables created
```
Questions from the above fit:

- What does each t-test do?
- What does $R^2$ mean?
- What does the F test do here? 




**`constrast()`**

To estimate the mean difference between two regions, we can use the `contrast()` function from package `contrast`.

```{r  results=TRUE, eval=FALSE}
contrast(fit.continent, list(Continent = 'As'), list(Continent = 'Am'))
contrast(fit.continent, list(Continent = 'As'), list(Continent = 'E'))  # comp As with E
```

**Changing base level**

We can choose the base level as we want. We demonstrate this by setting up European cars as the base category:

```{r results=TRUE}
data1$Continent <- factor(data1$Continent, levels = c("E", "Am", "As"))  
summary(lm(MPG_City ~ Continent, data1))
```

Note that:

1. The $F_{test}$ should be the same. 
2. The coefficients are different but the mean estimates are the same.


```{r}
# Change it back
data1$Continent <- factor(data1$Continent, levels = c("Am", "As", "E")) 
```

**Drawback of one way Anova:** other factors should be also taken into consideration.

## Model 2 - Model without Interaction

Let us take one important variable `Horsepower` in addition to the `Continent`. We first show three scatter plots by region with `lm` line added. Though the slopes seem to be little different for the sake of simple interpretation we  will fit a usual additive model without interaction, that is, we assume the effect of `HP` over `MPG_City` are the same. 

```{r}
ggplot(data1, aes(x = Horsepower, y = MPG_City, color=Continent)) + 
  geom_point() + 
  geom_smooth(method = "lm" , se =F) + 
  facet_wrap(~ Continent) + labs(title="Highway MPG by Weight")
# ggplot(data1, aes(x = Horsepower, y = MPG_City, color = Continent)) +
#   geom_point() +
#   geom_smooth(method = "lm" , se =F) +
#   labs(title = "Model With Interactions")
```

We next plot our models without interactions.
```{r}
fit.no.interation <- lm(MPG_City ~ Continent+Horsepower, data1)
coefs <- (summary(fit.no.interation))$coefficients[, 1]

ggplot(data1, aes(x = Horsepower, y = MPG_City, color = Continent)) + 
  geom_point() + 
  geom_abline(intercept =coefs[1], slope=coefs[4], color ="#F8766D") + 
  geom_abline(intercept =coefs[1]+coefs[2], slope=coefs[4], color = "#00BA38") +
  geom_abline(intercept =coefs[1]+coefs[3], slope=coefs[4], color ="#619CFF") +
  labs(title = "Model Without Interactions")
```


For a model without interaction, we assume that the effect of `HP` on `MPG_{City}` is the same regardless of which region the cars are produced.

$$MPG = \beta_{Am} + \beta_{As} \cdot I(Continent = As) + \beta_E \cdot I(Continent = E) + \beta_1 \cdot HP + \epsilon$$

```{r results=TRUE}
fit.no.interation <- lm(MPG_City ~ Horsepower + Continent, data1)
summary(fit.no.interation)   
```

To test whether `Continent` is significant after controlling for `Horsepower`, we can use `anova(H_0, H_1)`.

```{r results=TRUE}
anova(lm(MPG_City ~ Horsepower, data1), fit.no.interation  )    
```

p-value = 0.0237, a weak evidence of rejecting the null hypothesis that there is no `Continent` effects.


### `Anova()`

We can also use `Anova()` from the `car` package.

```{r results=TRUE}
Anova(fit.no.interation)    ### anova(fit.no.interation) # doesn't give us the right F test!!!! It is a sequencial test: given the variables above, do we need the current one?
```

We fail to reject $H_0$ at $\alpha=0.01$, i.e., controlling `HP`, we do not have the evidence to say the effect of `Continent` is different.


**Remark:** Models without interactions are simple and we can test Continent effects directly. But we may ask ourselves is this assumption reasonable?


## Model 3 - Model with Interaction

It is possible that  `HP` effects depend on  `Continent`.

```{r eval=T}
ggplot(data1, aes(x = Horsepower, y = MPG_City, color = Continent)) + 
  geom_point() + 
  geom_smooth(method = "lm" , se =F) + 
  labs(title = "Model With Interactions")
```

As we can see from the plot, Asian cars with smaller HP are more efficient; while European cars with larger HP are more efficient. To capture this effect, we introduce models with interactions `MPG` ~ `Horsepower` by `Continent`. 

$$y_{i|Am,\, HP} = \beta_{Am} + \beta_{1_{Am}} \cdot HP + \epsilon_i$$
$$y_{i|As,\, HP} = \beta_{As} + \beta_{1_{As}} \cdot HP + \epsilon_i$$

$$y_{i|E,\, HP} = \beta_{E} + \beta_{1_{E}} \cdot HP + \epsilon_i$$

This can be done through the following indicators:


$$MPG= \beta_{Am} + \beta_{As} \cdot I(Continent = As) + \beta_E \cdot I(Continent = E) + \\
 \beta_{1_{Am}} \cdot HP + \beta_{1_{As}} \cdot HP \cdot I(Continent = As) + \beta_{1_E} \cdot HP \cdot I(Continent = E) + \epsilon$$


Similarly, 

* $\beta_{As}$ is the increment of intercept between Asian and American, ... 

* $\beta_{1_{As}}$ is the increment of slope between Asian and American, ...

Use `lm()` with interaction: 

```{r results=TRUE}
fit.with.interaction <- lm(MPG_City ~ Horsepower*Continent, data1)
summary(fit.with.interaction)    #model.matrix(fit.with.interaction)[1:3,]
#summary(lm(MPG_City ~ Horsepower + Continent + Horsepower*Continent, data1)) # this is same as fit.with.interaction
```

Question: Is the interaction effect significant? Are the effects of HP the same across `Continent`? 

To test there is no interaction is same as to test
$$H_0: \beta_{1_{AS}} = \beta_{1_E} = 0$$
which we can do with `anova()`
```{r results=TRUE}
anova(fit.no.interation, fit.with.interaction)
Anova(fit.with.interaction)
```


We do reject the null hypothesis with a $p$-value to be $.0013$, not terribly small. That indicates some interaction effect of HP and Continent.

So the final conclusion would be

- Asian cars with smaller HP are more efficient; 
- European cars with larger HP are more efficient.

The interpretation of the model is no longer neat.

\pagebreak

# A more flexible model

We now try a model which includes all sensible variables.

```{r}
names(data1)
names.exclude <- names(data1) %in% c("Make.Model", "MPG_Hwy", "Make")
# names.exclude <- c("Make.Model", "MPG_Hwy", "Make") # doesn't work.
data2 <- data1[!names.exclude]  # Take a subset with all var's but "Make.Model","MPG_Hwy","Make"
names(data2)    #str(data2)   table(data2$Transmission)
```

Or use `dplyr`

```{r results=TRUE}
data2 <- select(data1, -Make.Model, -MPG_Hwy, -Make)
names(data2)
```

**QUESTION**: What would have happened if you have added Make.Model into the model???

```{r results=TRUE}
fit.all <- lm(MPG_City ~., data2)    #convenient way to include all variables
summary(fit.all)
data3 <- data2
data3$Cylinders <- as.factor(data3$Cylinders)
Anova(lm(MPG_City ~., data3) )   # interestingly Cyliners is very sig as a categorical variable.
```

The $t$-table is pretty messy, many variables are not significant, need to do `Anova()`

```{r results=TRUE}
Anova(fit.all)
```



Comments:

1. `Continent` is not needed after accounting for all other variables. That says we don't have evidence to show that one continent tends to produce more efficient cars than the others.
2. Can we drop all the variables with $p$-values larger than $.05$?
3. Take a look at `Transmission`. There are predominately automatic cars. Perhaps we should only include automatic cars and change our population. 

```{r results=TRUE}
table(data2$Transmission)
```
4. We don't have to use a model with all $p$-value being small!



**Model diagnosis**: check if three assumptions of linear models are met.

Residuals vs Fitted and qqnormal Plot, both plots look fine.
```{r  eval = T}
par(mfrow=c(1,2))
plot(fit.all, 1) 
plot(fit.all, 2) 
```


We now provide confidence Interval for the mean of cars like our `newcar`
```{r results=TRUE}
fit.mean.new <- predict(fit.all, newcar, interval = "confidence") 
fit.mean.new
```


We also provide prediction interval on this `newcar`
```{r results=TRUE}
fit.new <- predict(fit.all, newcar, interval = "prediction", se.fit=TRUE)
fit.new$fit
fit.new$se.fit
fit.new$residual.scale
```


# Summary

Summary about linear models:

   1. Linear model is simple and has nice interpretation
   
   2. Coefficient for each predictor depends on the model
    
   3. LS automatically adjusts the unit of predictors
   
   4. $t$-tests: the effect of each coeff. (Does not rank the importance of the predictors)
    
   5. $F$-tests: test a set of predictors (t-tests being special cases!)
   
   6. Model assumptions: all in residuals 
   
   7. The most damaging model violation is the presence of heteroscedasticity. It threatens the validity of each test. But we can use sandwich estimators to get the standard errors. See section Heteroscedasticity in the appendix.
   
   8. Last but not least: linear models used here only show associations between predictors and the response on average. We CANNOT conclude causality at all.


**BIGGEST QUESTION: How to identify a set of important variables or which model to use?????**

Choose models with small "prediction errors" is one way to go. We will discuss more in a lecture about model selection.







\pagebreak

# Appendices

##    Appendix 1: Clean dataset 

We show how we cleaned the dataset `Cars_04.csv` into `car_04_relular.csv`
```{r eval=F}
# Caution: you need to have Cars_04.csv in order to run this chunk!!!!!
data = read.csv("data/Cars_04.csv", header=T, as.is=F)
str(data)   # There are 242 cars in the data
data.comp <- na.omit(data)  # It deletes any car with a missing value - not a good idea.
str(data.comp)  #182 cars with complete records for all the variables. 

## Explore the data
names(data)
head(data)
summary(data)

par(mgp=c(1.8,.5,0), mar=c(3,3,2,1)) 
hist(data$Horsepower, breaks=20, col="blue")  # notice that there are some super powerful cars

## Let's find out which are the super powerful cars
data$Make.Model[data$Horsepower > 400] # Show models with Horsepower > 400
data$Make.Model[data$Horsepower > 390]
data$Make.Model[data$Horsepower <= 100]

## We could find cars with horsepower > 400
data[data$Horsepower > 400, "Make.Model"] 

## Let's concentrate on  "regular" cars and exclude those super expensive ones 
datacar <- data[(data$Horsepower <=390) & (data$Horsepower > 100), ]

## Take a subset with relevant variables
variables <-  c("Make.Model", "Continent", "MPG_City", "MPG_Hwy",
              "Horsepower","Weight","Length", "Width",
              "Seating","Cylinders","Displacement",
              "Make", "Transmission")

data1 <- datacar[, variables]  # subset
str(data1)
write.csv(data1, file="car_04_regular.csv", row.names = FALSE)
                            # Output the data and name it car_04_regular.csv

```


##    Appendix 2: Using formulae to obtain OLS estimates

Take `fit3` and recreate $\hat\beta$'s. 
```{r}
design.x <-  model.matrix(fit3)  #fit3$model gives us Y and x1 x2...xp 
y <- fit3$model[, 1]
design.x[, 1]
rse <- (summary(fit3))$sigma
beta <-  (solve( t(design.x) %*% design.x)) %*% t(design.x) %*% y # reconstruct LS estimates
beta
```

Reconstructing the covariance matrix from `fit3`
```{r}
summary(fit3)$cov.unscaled   # inverse (X' X)
cov.beta <- (rse^2) * (summary(fit3)$cov.unscaled)  # alternatively we can get the cov matrix this way
```


Calculating the Standard Error from `fit3`
```{r}
sd.beta <- sqrt(diag(cov.beta)) # check to see this agrees with the sd for each betahat
sd.beta  # this shoulbe be same as the Std. Error in the summary table

summary(fit3)$coeff[, "Std. Error" ]
```

## Appendix 3: Heteroscedasticity

**Sandwich estimator**:

In a linear model, when $\text{Var}(y_i| x_i)=\sigma^2(x_i)$, i.e., the linear model is hereroscedastic. While the least squared estimator of $\beta$ is still unbiased but, the usual variance $\text{Var}( {\bf\hat{\beta}}) \neq \sigma^2 (X^T X)^{-1}$. It has been long known among in econometrics that we can correct the variance estimator by a `Sandwich estimator`. Since
$$\text{Var}( {\bf\hat{\beta}}) = (X^T X)^{-1}(X^T D X) (X^T X)^{-1}$$
where $D$ is a diagonal matrix of $d_i=\sigma^2_i=\text{Var}(y_i|x_i)$. We of course do not know $d_i$. So we use squared residual to estimate $d_i$: $\hat d_i = (y_i-\hat y_i)^2$. That yields famous Sandwich Estimator due to White (1980): 
$$\text{Var}( {\bf\hat{\beta}}) = (X^T X)^{-1}(X^T \hat D X) (X^T X)^{-1}$$
This correction has been implemented in various packages. We use `sandwich` with function `lmtest::coeftest(fm, df = Inf, vcov = vcovHC)`. `HC` means:Heteroscedasticity-Consistent Covariance Matrix Estimation. 

To produce a HC corrected confidence intervals for $\beta$, we first get a linear fit as  $\text{fit} = lm(y~....)$. Then apply: `lmtest::coeftest`(fit, df = Inf, vcov = vcovHC) to get the correct confidence intervals for $\beta$. 

**Remark:** Even linearity is not there, we can still fit a linear model. We can interpret the linear function as an approximation to the mean. The confidence intervals obtained using the sandwich estimators are still valid. See our recent work *Models as Approximations I: Consequences Illustrated with Linear Regression*. 

Let us revisit the final model with all variables:
```{r results=TRUE}
data2 <- select(data1, -Make.Model, -MPG_Hwy, -Make)
fit.all <- lm(MPG_City ~., data2) 
Anova(fit.all)
plot(fit.all, 1)
```
From the residual plot, one may worry the violation of linearity, as well as presence of heteroscedasticity. 

To provide a more trust-worthy analysis we may use sandwich standard errors for a more accurate inferences over coefficients.  


```{r results = TRUE}
#sandwich(fit.all)  # gives us HC cov estimator
#vcovHC(fit.all) # also gives us the HC cov matrix
fit.all.hc <- lmtest::coeftest(fit.all, df = Inf, vcov = vcovHC)
fit.all.hc  # output
```

Let compare regulare lm output and the one with sandwich cov estimation:
```{r results = "asis"}
stargazer(fit.all, fit.all.hc, type = output_format ) 
```
While the sandwich standard errors remain more or less the same as before, we do notice one difference: Asian cars are more efficient than that of American cars at $\alpha = .1$. 

The estimators for each coefficient remain the same!

```{r}
cov <- vcovHC(fit.all, type = "HC")
robust.se <- sqrt(diag(cov))
stargazer(fit.all, fit.all.hc, type = output_format,
          se = list(NULL, robust.se)) 
```

**Remark:** How to incoporate heteroscedasticity in prediction intervals? One easy way out is to fit another model taking residual square as response and use predicted values as the variance give $x$. We could use a more relaxed non-linear model such as trees/Random Forests. 




##  Appendix 4: Nonlinearity and transformations

The $x$ variable may relate to $y$ through a function of $x$, say $log(x)$ or $x^{2}$. 

  * We can handle this by transforming $x$.
  * First let's find out if there is some non-linear relationship there by looking at pairwise scatter plots


Based on the following `Horsepower` may need a transformation
```{r eval=T}
pairs(data1[, -1]) 
```


We look at `Horsepower` vs. $\frac{1}{Horsepower}$
```{r message = F, eval=T}
par(mfrow=c(1,2))
plot(data1$Horsepower, data1$MPG_City, pch=16) 
plot(1/data1$Horsepower, data1$MPG_City, pch=16)  #looks better!
```


$\frac{1}{Horsepower}$ Looks good, so we make a model using that.
```{r}
fit.transform <- lm(MPG_City~ I(1/Horsepower), data1)  # Use I()
plot(fit.transform, 1)
```

We also try to fit $Horsepower^{2}$
```{r}
fit12 <- lm(MPG_City ~ Horsepower + I(Horsepower^2), data1) # fit a quadratic function
summary(fit12)
plot(fit12$fitted, fit12$residuals, pch=16)
abline(h=0, lwd=4, col="red")
plot(fit12, 2)
```






## Appendix 5: Model Diagnoses via a perfect linear model

We look to fit a model containing 100 points of the equation 

$$y = 1 +2x + \mathcal{N}(0,2) \quad i.e. \quad \beta_0 = 1 ,\beta_1 = 2, \sigma^2 = 2.$$

```{r}
par(mfrow=c(1,1))

x <- runif(100)
y <- 1+2*x+rnorm(100,0, 2)
fit <- lm(y~x)
fit.perfect <- summary(lm(y~x))
rsquared <- round(fit.perfect$r.squared,2)
hat_beta_0 <- round(fit.perfect$coefficients[1], 2)
hat_beta_1 <- round(fit.perfect$coefficients[2], 2)
plot(x, y, pch=16, 
     ylim=c(-8,8),
     xlab="a perfect linear model: true mean: y=1+2x in blue, LS in red",
     main=paste("R squared= ",rsquared, 
                ", LS estimates b1=",hat_beta_1, "and b0=", hat_beta_0))
abline(fit, lwd=4, col="red")
lines(x, 1+2*x, lwd=4, col="blue")
```


Residual plot
```{r}
plot(fit$fitted, fit$residuals, pch=16,
     ylim=c(-8, 8),
     main="residual plot")
abline(h=0, lwd=4, col="red")
```




Check normality
```{r}
qqnorm(fit$residuals, ylim=c(-8, 8))
qqline(fit$residuals, lwd=4, col="blue")
```





The following example tells us we can't look at y directly to check the normality assumption. Why not? We really need to examine residuals!

```{r}
par(mfrow=c(2,2))
x <- runif(1000)
y <- 1+20*x+rnorm(1000,0, 2)
plot(x, y, pch=16)
hist(y, breaks=40)
qqnorm(y, main="qqnorm for y")
qqline(y)
fit <- lm(y~x)
qqnorm(fit$residuals, main="qqnorm for residuals")
qqline(fit$residuals)
par(mfrow=c(1,1))

data.frame( mean = mean(y), std.dev = sd(y))
```


## Appendix 5: Colinearity

When some $x$'s are highly correlated we can not separate the effect. But it is still fine for the purpose of prediction.

A simulation to illustrate some consequences of colinearity. Each $p$-value for $x_1$ and $x_2$ is large but the null of both $\beta's = 0$ are rejected....  Because $x_1$ and $x_2$ are highly correlated.

```{r}
par(mfrow=c(2,1))
set.seed(1)  
x1 <- runif(100)
x2 <- 2*x1+rnorm(100, 0,.1)    # x1 and x2 are highly correlated
y <- 1+2*x1+rnorm(100,0, .7)   # model
newdata.cor <- cbind(y, x1,x2) # to see the strong correlations..
pairs(newdata.cor, pch=16)
```

$x_1$ is a useful predictor
```{r results=TRUE}
summary(lm(y~x1)) 
```

$x_2$ is a useful predictor
```{r results=TRUE}
summary(lm(y~x2))
```

$x_1$ and $x_2$ together show that one is not
```{r results=TRUE}
summary(lm(y~x1+x2)) 
```

Putting both highly correlated var's together we can't separate the effect of each one, though the model is still useful!
```{r results=TRUE}
data.frame(Corealation = cor(x1, x2))
```
Model dignoses show validities of model assupmtions. 
```{r}
plot(lm(y~x1+x2), 1:2)
```

##    Appendix 6: F-distribution
A quick look at an $F$ distribution. One may change $df_1 = 4$ and $df_2=221$ to see the changes in the distribution.
```{r  eval = T}
## 
hist(rf(10000, 6, 226-7), freq=FALSE, breaks=200)   # pretty skewed to the right

Fstat=(summary(fit3))$fstat    # The Fstat, df1, df2
pvalue=1-pf(Fstat[1], 6, 219) #or pf(Fstat[1], 6, 219, lower.tail=FALSE)

 # As long as Fstat is larger than 2.14, we reject the null at .05 level.
data.frame(F_stat = Fstat[1] , pvalue = pvalue , Cutoff = qf(.95, 6, 219))
```
