title: "WDS, HW 2"
author:
- Aaron Tom
- Neil Song
- Ben Zhao
- Luke Melcher
- Tanvi Kigga
date: 'Due: 10:00PM, Friday, July 14, 2021'
output:
  html_document:
    code_folding: show
    highlight: haddock
    theme: lumen
    toc: yes
    toc_depth: 4
    toc_float: yes
  pdf_document:
    toc: yes
    toc_depth: '4'
urlcolor: blue
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, results = "hide", fig.width=8, fig.height=4)
options(scipen = 0, digits = 3)  # controls base R output
# check if you have ISLR package, if not, install it
if(!require('pacman')) {install.packages('pacman')}
pacman::p_load(ISLR, readxl, tidyverse, magrittr, dplyr, ggplot2)
```


\pagebreak


# Objectives 

*The main purpose for this homework is to tell the difference between probability and statistics.* Using the Roulette game as illustration, we go through the main elements of probability and statstics. We then learn how to estimate unknown parameters in a population with precision, i.e., a confidence interval. Simulations help us to understand the interpretation of a confidence interval. 

- Continue to get used to `RStudio` and `RMarkdown`
- Probability
  - Random variables and their distributions
  - Population parameters: expected values and variances
  - Normal variables
  - Central limit theorem (CLT)

- Statistics
  - Sample means and sample variances
  - Use sample mean to estimate the population mean
  - Confidence intervals/hypothesis tests
  
- Packages/functions
    - `dplyr`
    - `ggplot`
    - `pnorm()`, `qnorm()`, `dnorm()`


# Review materials


- Study Module 2: Probability_Statistics_101


# Question 1: Probability

The following questions refer to the Red-Black Roulette game. After the ball is spun it will land in one of the slots.

a.	What is the probability that the ball will land in any one slot?

The probability a ball will land on a specific spot is a 1 in 38 chance. 

b.	A player will bet \$1 each time. Let $X$ be the payoff.

    (i)	If the player bets on a single number, write down the distribution of $X$, i.e. the possible values of $X$ and the corresponding probabilities. Calculate the expected value of winning and the standard deviation.
    
If a player bets on a single number, the distribution of $x$ is a binomial distribution, would be 1/38 with a possible value of +1 for the player and 37/38 for the casino with a possible value of -1.
```{r single_bet}
  expected_1 <- 1/38*35 + 37/38*-1
  variance_1 <- (35-(-0.0526))^2 * 1/38 + (-1 -(-0.0526))^2 * 37/38
  standard_1 <- sqrt(variance_1)
```
      
    
    (ii)	Repeat number (i) if the player bets on two numbers.
```{r double_bet}
  expected_2 <- 2/38*35 + 36/38*-1
  variance_2 <- (35-(-0.0526))^2 * 2/38 + (-1 -(-0.0526))^2 * 36/38
  standard_2 <- sqrt(variance_2)
```
    
    (iii)	If one plays the game by betting on a single number 1000 times, how much do you expect the person to win?
```{r}
# expected gain


n <- 1000
win_prob <- 1/38
# winning event
set.seed(2021)
win_vec <- rbinom(n,1,win_prob) 
# if win: +35; if lose: -1
gain_vec <- win_vec*36
gain_vec <- gain_vec-1
# print first 100 gains
# head(gain_vec, 100)
# to find average gain
sum(gain_vec)
```
  According to our seed simulation, we would expect the person to win $8.
    
    (iv)	Explain now why the casinos always make money on the roulette game.
    
Casinos always make money on the roulette game because no matter what way you position yourself- all black, all red, or a specific number, the casino is favored to win. The highest probability you could give yourself is 18/38, and that in itself is not above 50%, making the results biased towards the casino. Taking into account the concept of expected value, as the amount of games increases, the more likely you are to approach that value- thus the casinos are extremely likely make money on the roulette game as the number of trials approach larger and larger numbers.

c. Let's "play" the game using code.  

    (i) We bet 30 times each round. Put 30 payoffs into a vector and store into a variable.
```{r}
set.seed(1)
num_win = rbinom(30, 30, 1/36)
num_win
```

   (ii) Plot the histogram of the 30 payoffs. What is the total payoffs for this round?
```{r}
num_lose <- 30- num_win
avg_gain <-  (num_win - num_lose)/30
hist(avg_gain)

```