---
title: "Data Acquisition, Preparation and EDA"
author: 
output:
  beamer_presentation:
    slide_level: 3
    theme: Boadilla
    colortheme: rose
  ioslides_presentation: default
  slidy_presentation: default
header-includes:
- \AtBeginSection[]{\begin{frame}\tableofcontents[currentsection]\end{frame}}
- \AtBeginSubsection[]{\begin{frame}\tableofcontents[currentsubsection]\end{frame}{}}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, cache = T, fig.width=8, fig.height=4)
options(scipen = 0, digits = 3)  # controls base R output
if(!require('pacman')) {
  install.packages('pacman')
}
pacman::p_load(ggplot2, dplyr, tidyverse, gridExtra, ggrepel, plotly, skimr, tidytext) 
# install a package if it does not exist already and put the package in the path (library)
# dplyr, ggplot2,tidyr
```

```{r echo = F, eval = F}
# To properly generate PDF slides
# you need to install latex as follows
install.packages("tinytex")
tinytex::install_tinytex()  # install TinyTeX
```


# Introduction and Objectives

- **Data Science** connects statistics, computer science, and domain knowledge. 
- We look for patterns & reasons for differences/changes in datasets.


---
\frametitle{Introduction and Objectives}

- Datasets
  - design a study
    - lay out goals
    - apply domain knowledge (what factors are important?)
    - generate variable list
    - gather data (experiments, surveys, other studies)
    - account for study cost and feasibility
  - OR analyze existing data


---
\frametitle{Introduction and Objectives}

- Once we have the data, we proceed to extract useful information
- BUT we must understand the data first

- In this lecture:
  - basic data acquisition/preparation
  - understand the nature of the data  
  via **exploratory data analysis (EDA)**
  - explore plausible variable relationships
- *We defer formal modeling for later*


---
\frametitle{Introduction and Objectives}

- Data mining tools:
  - expanding dramatically in the past 20 yrs
- R
  - popular among data scientists & in academia
  - open-source
  - most SOTA methods have R package implementations


---
\frametitle{Introduction and Objectives}

- The number of studies is soaring (especially during COVID)
- BUT a significant number cannot be reproduced/replicated
  - this phenomenon is an ongoing crisis termed the **replicability crisis**
- In an effort to produce trustworthy and reproducible results,  
we use **R Markdown**.
  - It achieves many goals:
    1) Anyone can rerun our study to replicate the results.
    2) We can run our data analysis & produce reports at the same time.
- Communication between us and readers/decision-makers is essential.


---
\frametitle{Objectives}


- This module will focus on **data preparation**, **data cleaning**,  
and **exploratory data analysis (EDA)**.  
- R and R Markdown will be used.  
See advanced_R_tutorial.Rmd for extremely useful EDA tools  
such as `dplyr`, `ggplot`, `data.table`, and more.


---
\frametitle{Contents}

0. Suggested extra readings/doing:  
    + run and study Get_staRted.Rmd
    + run and study advanced_R_tutorial.Rmd and advanced_R_tutorial.html
    + read `50 years of data science` and `Teaching data science` available in Canvas. (skip this)
    + Data set: `MLPayData_Total.csv`
1. Case Study: Billion dollar Billy Beane
2. Study flow:
    + Study design
    + Gathering data
    + Process data (tidy data)
    + Exploratory Data Analysis (EDA)
    + Conclusion/Challenges
4. R functions 
    + basic r functions
    + dplyr
    + ggplot
    

---
\frametitle{Handy Cheat Sheets}

**[DPLYR Cheat Sheet](http://www.rstudio.com/wp-content/uploads/2015/02/data-wrangling-cheatsheet.pdf):**  
http://www.rstudio.com/wp-content/uploads/2015/02/data-wrangling-cheatsheet.pdf

**[ggplot Cheat Sheet](https://github.com/rstudio/cheatsheets/blob/master/data-visualization-2.1.pdf):**  
https://github.com/rstudio/cheatsheets/blob/master/data-visualization-2.1.pdf

# Case Study: Baseball  
**Background**  

- Baseball is one of the most popular sports in the US. 
- Major League Baseball (MLB) includes the highest level of baseball teams
  - 30 total teams including the American League and the National League
  - top teams include the New York Yankees, Boston Red Sox, Philadelphia Phillies, and Oakland Athletics
- Oakland A's
  - low budget team
  - recent rising star
  - General Manager (GM) Billy Beane  
  is well known to apply statistics in his coaching


---
\frametitle{Case Study: Baseball}

**Background**

- In the article [Billion Dollar Billy Beane](https://fivethirtyeight.com/features/billion-dollar-billy-beane/):
  - author: Benjamin Morris
  - studies a regression of performance vs. total payroll  
  from all 30 teams over a 17 yr period
  - Oakland A's performance is comparable to the Boston Red Sox
  - therefore argues that Billy Beane is worth $12 million over 5 years
- Article link: [https://fivethirtyeight.com/features/billion-dollar-billy-beane/](https://fivethirtyeight.com/features/billion-dollar-billy-beane/)


---
\frametitle{Case Study: Baseball}

**Objectives**  

1. Reproduce Benjamin's study. 
  - Is Billy Beane (Oakland A's GM) worth 12.5 million dollars for a period of 5 years, as argued in the article? 
  - We challenge Benjamin's reasoning behind his argument. 

2. Explore general questions: 
  - How does pay and performance relate to each other? 
  - Will a team perform better when they are paid more?


---
\frametitle{Gathering Data}


- We gathered information on payroll and performance by team from 1998 to 2014 and reassembled the data set from several websites. 
  - (We could have easily *reproduced* all the analyses from the post if the article's data was available!)
- Some manual corrections were made. 
  - i.e. consolidating team names for teams that underwent a name change.
- Example source: http://www.stevetheump.com/Payrolls.htm


---
\frametitle{Gathering Data}

**Data:** `MLPayData_Total.csv`, consists of winning records and the payroll of all 30 ML teams from 1998 to 2014 (17 years). There are 162 games in each season.

The variables included are:

* `team name`: team names
* `p2014`: total pay in 2014 in **millions** and other years indicated by year
* `X2014`: number of games won in 2014 and other years labeled
* `X2014.pct`: percent winning in 2014 and other years (We only need one of the two variables from above.)


## Data Preparation

---
\frametitle{Data Preparation}

\begin{alertblock}{}
Before we do any analysis, it is a \textbf{MUST} that we take a look at the data.
\end{alertblock}
In particular, we will try to: 

**Tidy the data:**

+ Import data/Data preparation
+ Data format
+ Missing values/peculiarity
+ Understand the variables: unit, format, unusual values, etc.
+ Put data into a standard data format

Columns: variables  
Rows: subjects


---
\frametitle{Data Preparation}

**Read the data:**

The simplest way to import data is to use **`read.csv()`**  
(or the faster `fread()` from the `data.table` package).  

What is the current working directory?  
R will find files or save files to the working directory.  
\small
```{r eval=F}
getwd()
dir <- "/Users/lzhao/Dropbox/STAT471/Data"  # my laptop
setwd(dir)   #same as  setwd("/Users/lzhao/Dropbox/STAT471/Data")
getwd()
```
\normalsize
Alternatively, if the data is in the same folder as the .Rmd file then we can read data directly.  

\small
```{r read data}
datapay <- read.csv("data/MLPayData_Total.csv", header=T, stringsAsFactors = FALSE)
```


---
\frametitle{Data Preparation}

**What is in the dataset?**  
Take a quick look at the data. Pay attention to what is in the data, any missing values, and the variable format.   
\tiny
```{r, exam variable name}
names(datapay)
```
\normalsize
Is anything bothering you? We may want to change names of teams to a shorter, neater name.


---
\frametitle{Data Preparation}

Everything seems to be OK at the moment other than changing one variable name.  
\tiny
```{r, exam data} 
#summary(datapay)
summary(datapay)[1:10] # quick summary. missing values may be shown
str(datapay) # data structure
```

---
\frametitle{Data Preparation}

Let's update the team name variable.  
\tiny
```{r change name}
# change variable name and also update the data file
datapay <- datapay %>% rename(team = Team.name.2014) 
names(datapay)[1:5] # only show 5 names
```


---
\frametitle{Data Preparation: Reshape the data}


The original format of the dataset `MLPayData_Total.csv` is not in a desirable format. Each row lists multiple results. Also the variable `year` is missing.  
\tiny
```{r exam a few lines and columns}
datapay[1:4, 1:5]  # list a few lines (subsetting) 
#datapay$team # get variables
```


---
\frametitle{Data Preparation: Reshape the data}


We would like to reshape the data into the following table format:  

- columns (variables) contain all variables  
- each row records one result(s)   

In our case we have four variables: team, year, pay, win_number and win_percentage. Let's rearrange the data into the following form:  

team | year | payroll | win_number | win_percentage 


---
\frametitle{Data Preparation: Reshape the data}


Let us do this using `dplyr::pivot_longer()`.  
First we create the `payroll` and `year` variables:  
\tiny
```{r, long table, results="hide"}
payroll <- datapay %>%   # first create variable: payroll and year
  select(team, p1998:p2014) %>% 
  pivot_longer(cols = starts_with("p"), 
               names_to = "year", 
               names_prefix = "p",
               values_to = "payroll")
payroll[1:3, 1:3] # show a few rows
```


---
\frametitle{Data Preparation: Reshape the data }


Let's create the other variables.
\tiny
```{r, results="hide"}
win_num <- datapay %>%  # create variable: win_num and year
  select(team, X1998:X2014) %>% 
  pivot_longer(cols = X1998:X2014,
               names_to = "year", 
               names_prefix = "X",
               values_to = "win_num")

win_pct <- datapay %>%  # create variable: win_pct and year
  select(team, X1998.pct:X2014.pct) %>% 
  pivot_longer(cols = X1998.pct:X2014.pct,
               names_to = "year",
               names_prefix = "X", 
               values_to = "win_pct") %>%
  mutate(year = substr(year, 1, 4))
```


---
\frametitle{Data Preparation: Reshape the data  }


Finally, we join the tables into team, year, payroll, win_num, and win_pct.  
\tiny
```{r}
datapay_long <- payroll %>% 
  inner_join(win_num, by = c("team", "year")) %>%
  inner_join(win_pct, by = c("team", "year")) 
head(datapay_long, 2)  # see first 2 rows
```


---
\frametitle{Data Preparation: Reshape the data  }


Take a quick look at the newly formed data file `datapay_long`. 
\tiny
```{r}
names(datapay_long) #names(datapay) new vs. old data files
```
\normalsize
Quick summary of the new data:
\tiny
```{r, summary long table}
head(datapay_long) # shows the first 6 rows
```
\normalsize
More ways to summarize the data:
\tiny
```{r}
# dim(datapay_long) 
# str(datapay_long)
# summary(datapay_long)
# skimr::skim(datapay_long)
```


---
\frametitle{Data Preparation: Output the cleaned data file   }


After processing, save this cleaned data file into a new table called `baseball.csv`.
Let's output this table to the /data folder in our working folder. From now on we will only use the data file `baseball`.
\tiny
```{r, store a file}
write.csv(datapay_long, "data/baseball.csv", row.names = F)
```
\normalsize
\begin{block}{Remark}The above data prep process should be put into a separate .r or .rmd file. There is no need to rerun the above data prep portion each time we work on the project. We put the whole project into one file for the purpose of demonstration.
\end{block}

# Exploratory Data Analysis (EDA)
- All the analyses done in this lecture will be exploratory. 
- The goal is to \alert{see what information we might be able to extract} so that it will support the goal of our study. This is an extremely important first step of the data analyses. 
- We try to understand the data, summarize the data, then finally explore the relationships among the variables through useful visualization.

## Part I: Analyze aggregated variables  

In this section:  

  - Try to use aggregated information such as:
    - the total pay for each team 
    - average performance
  - Look for the relationship between performance and the payroll as suggested in Morris's post. 


---
\frametitle{Input the data}


First, input the clean data `baseball` and quickly explore the data.  
Everything seems fine: no missing values, names of variables are good. The class of each variable matches its nature. (numeric, factor, characters...)  
\tiny
```{r, read baseball data, results= "hold"}
baseball <- read.csv("data/baseball.csv", header = TRUE, stringsAsFactors = F)
names(baseball)
str(baseball)
summary(baseball)
#View(baseball)
```


---
\frametitle{Create a new data table}

For convenience, we create a new table which only contains the total payroll and average winning percentage for each team. We name them  `team`, `payroll_total`, and `win_pct_ave`. We will change the unit of `payroll_total` from million to billion.  
\tiny
```{r, aggeragate w summarise, results="hold"}
# create total and average winning percentage for each team
data_agg <-baseball %>% 
  group_by(team) %>%
  summarise(
    payroll_total = sum(payroll)/1000, 
    win_pct_ave = mean(win_pct))
str(data_agg)
summary(data_agg)
```


---
\frametitle{Descriptive statistics}

To summarize a continuous variable (such as `payroll_total` or `win_pct_ave`), we use the following measurements:  

+ **Center**: sample mean/median
+ **Spread**: sample standard deviation
+ **Range**: minimum and maximum 
+ **Distribution**: quantiles


---
\frametitle{Descriptive statistics}

First, let us take a look at `payroll_total`.   

**Base `R` way:**
\tiny
```{r, payroll_total summary 1, results = "hold"}
mean(data_agg$payroll_total)
sd(data_agg$payroll_total) 
quantile(data_agg$payroll_total, prob = seq(0, 1, 0.25))
median(data_agg$payroll_total)
max(data_agg$payroll_total)
min(data_agg$payroll_total)
summary(data_agg$payroll_total)
```


---
\frametitle{Descriptive statistics}

First, let us take a look at `payroll_total`.  

**`dplyr` way:**
\tiny
```{r payroll_total summary 2, results="hold"}
data_agg %>% select(payroll_total) %>% 
  summarise(
    mean = mean(payroll_total),
    sd   = sd(payroll_total),
    max = max(payroll_total),
    min = min(payroll_total),
    "0%" = quantile(payroll_total)[1],
    "25%" = quantile(payroll_total)[2],
    "50%" = quantile(payroll_total)[3],
    "75%" = quantile(payroll_total)[4],
    "100%" = quantile(payroll_total)[5]
  )
```


---
\frametitle{Descriptive statistics}

Find the team with the max/min payroll.  

**Base `R` way:**  
\tiny
```{r}
data_agg$team[which.max(data_agg$payroll_total)] 
data_agg$team[which.min(data_agg$payroll_total)]
```


---
\frametitle{Descriptive statistics}

Rearrange the data to see the ranks of team by `payroll`.  

But we can easily rearrange the whole data set `data_agg` by ordering one variable, say `payroll_total`.  

**Base `R` way:**
\tiny
```{r}
#To rank teams by payroll in decreasing order
arrange(data_agg, desc(payroll_total))[1:5,] #default decs=T
#arrange(data_agg, win_pct_ave) # default???
#arrange(data_agg, -desc(payroll_total))[1:5,] 
```


---
\frametitle{Descriptive statistics}

Rearrange the data to see the ranks of team by `payroll`.  

\normalsize
**`dplyr` way:**
\tiny
```{r results="hold"}
data_agg %>% select(team,payroll_total) %>% filter(payroll_total == max(payroll_total))
data_agg %>% select(team,payroll_total) %>% filter(payroll_total == min(payroll_total))
```

\normalsize
**`dplyr` way:**
\tiny
```{r, arrange dplyr}
data_agg %>%
  arrange(payroll_total) %>%
  slice(1:5)  # select first 5 rows
```


---
\frametitle{Displaying variables: Histograms  }


**`ggplot` plots:**
\tiny
```{r, facet hist, fig.height = 2.5}
p1 <- ggplot(data_agg) + 
  geom_histogram(aes(x = payroll_total), bins = 10, fill = "blue") +
  labs( title = "Histogram of Payroll (frequency)", x = "Payroll" , y = "Frequency")

p2 <- ggplot(data_agg) + 
  geom_histogram(aes(x = payroll_total, y = ..density..), bins = 10, fill = "light blue") +
  labs( title = "Histogram of Payroll (percentage)", x = "Payroll" , y = "Percentage")

grid.arrange(p1, p2, ncol = 2) # facet the two plots  side by side 
```

\normalsize
Notice, the two plots above look identical but with different `y-scale`. 


---
\frametitle{Displaying variables: Boxplots}

A **boxplot** captures the spread by showing median, quantiles and outliers:

**`ggplot` plots:**  
\tiny
```{r boxplot, fig.height=2, fig.width=3}
ggplot(data_agg) + 
  geom_boxplot(aes(x="", y=payroll_total)) + 
  labs(title="Boxplot of Pay Total", x="") 
```


---
\frametitle{Normal variables}


When would the sample mean and sample standard deviation help us to describe the distribution of a variable? As an exercise, let us summarize the variable `win_pct_ave`.

\tiny
```{r results="hold"}
mean(data_agg$win_pct_ave) # sort(data_agg$win_pct_ave)
sd(data_agg$win_pct_ave)
```
\normalsize
We see that  `win` on average is `r mean(data_agg$win_pct_ave)` with a SD being `r sd(data_agg$win_pct_ave)`.
How would the mean and sd be useful in describing the distribution of `win`?
**Only if the histogram looks like a bell curve**!


---
\frametitle{Normal variables}

Take a look at the histogram of `win`. Here we impose a **normal curve** with the center being `r mean(data_agg$win_pct_ave)` and the spread, sd = `r sd(data_agg$win_pct_ave)`.

**`ggplot` way:**
\tiny
```{r, hist w normal, fig.height=2, fig.width=3}
ggplot(data_agg) +
  geom_histogram(aes(x=win_pct_ave, y = ..density..), bins=10, fill= "blue" ) +
  stat_function(fun = dnorm, args = list(mean = mean(data_agg$win_pct_ave), 
                                         sd = sd(data_agg$win_pct_ave)), colour = "red",                                            size = 1.5)+
  labs( title = "Histogram of win_pct_ave", x = "win_pct_ave" , y = "Frequency")
```

---
\frametitle{Normal variables}

The smoothed normal curve captures the shape of the histogram of win. Or we will say that the variable `win` follows a normal distribution approximately. Then we can describe the distribution of `win` using the two numbers: mean and sd.  

Roughly speaking:  

* 68% of teams with `win` to be within one sd from the mean. 
$$ `r mean(data_agg$win_pct_ave)` \pm `r sd(data_agg$win_pct_ave)`= [`r mean(data_agg$win_pct_ave) -  sd(data_agg$win_pct_ave)`, 
`r mean(data_agg$win_pct_ave) +  sd(data_agg$win_pct_ave)`]$$

* 95% of the teams with `win` to be within 2 sd from the mean:
$$ `r mean(data_agg$win_pct_ave)` \pm 2 * `r sd(data_agg$win_pct_ave)`= [`r mean(data_agg$win_pct_ave) -  2 * sd(data_agg$win_pct_ave)`, 
`r mean(data_agg$win_pct_ave) +  2 * sd(data_agg$win_pct_ave)`]$$

* 2.5% of the teams with `win` to be higher 2.5 times of sd above the mean:
$$ > `r mean(data_agg$win_pct_ave)` + 2 * `r sd(data_agg$win_pct_ave)`= 
`r mean(data_agg$win_pct_ave) +  2 * sd(data_agg$win_pct_ave)`$$


---
\frametitle{Explore variable relationships}


- **Scatter plots** show the relationship between $x$ variable `payroll_total` and $y$ variable `win_pct_ave`.  
- We are looking for patterns between the two variables, such as a linear or quadratic relationship.


---
\frametitle{Explore variable relationships: Scatter plots  }

**`ggplot` plots** (shown on next slide)
\tiny
```{r, scatter p with team names, fig.show='hide', warning=FALSE}
data_agg %>%
  ggplot(aes(x = payroll_total, y = win_pct_ave)) + 
  # geometric options: color, size, shape, alpha: transparency (range: 0 to 1)
  geom_point(color = "blue", size= 3, alpha = .8) + 
  geom_text_repel(aes(label = team), size = 3) +
  labs(title = "MLB Team's Overall Win  vs. Payroll", 
       x = "Payroll_total", 
       y = "Win_pct_ave")
```


---
\frametitle{Explore variable relationships: Scatter plots  }


**We notice the positive association:**  
**when `payroll_total` increases, so does `win_pct_ave`.** 

```{r, scatter p with team names 2, echo=FALSE, fig.height = 4.5, warning=FALSE}
data_agg %>%
  ggplot(aes(x = payroll_total, y = win_pct_ave)) + 
  # geometric options: color, size, shape, alpha: transparency (range: 0 to 1)
  geom_point(color = "blue", size= 3, alpha = .8) + 
  geom_text_repel(aes(label = team), size = 3) +
  labs(title = "MLB Team's Overall Win  vs. Payroll", 
       x = "Payroll_total", 
       y = "Win_pct_ave")
```


---
\frametitle{Explore variable relationships: Scatter plots}


We can bring in other variables to adjust the color, size, and alpha of the scatter plot via **aesthetic mapping**.

\tiny
```{r, scatter p with team names with mappings, fig.show = 'hide', warning=FALSE}
data_agg %>%
  ggplot(aes(x = payroll_total, y = win_pct_ave)) + 
  # geometric options with aes mapping: 
  # color, size, alpha as a function of a variable 
  geom_point(aes(color = team), size = 3) + 
  geom_text_repel(aes(color = team, label = team, size = payroll_total)) +
  labs(title = "MLB Team's Overall Win  vs. Payroll", 
       x = "Payroll_total", 
       y = "Win_pct_ave") +
  theme_bw() +
  theme(legend.position = "none")
```

---
\frametitle{Explore variable relationships: Scatter plots}


We can bring in other variables to adjust the color, size, and alpha of the scatter plot via **aesthetic mapping**.

```{r, scatter p with team names with mappings 2, echo=FALSE, warning=FALSE}
data_agg %>%
  ggplot(aes(x = payroll_total, y = win_pct_ave)) + 
  # geometric options with aes mapping: 
  # color, size, alpha as a function of a variable 
  geom_point(aes(color = team), size = 3) + 
  geom_text_repel(aes(color = team, label = team, size = payroll_total)) +
  labs(title = "MLB Team's Overall Win  vs. Payroll", 
       x = "Payroll_total", 
       y = "Win_pct_ave") +
  theme_bw() +
  theme(legend.position = "none")
```


---
\frametitle{Explore variable relationships}

**Least Squared Lines**  

- The simplest function to capture the relationship between pay and performance is through the linear model.  
- We impose the least squared equation on top of the scatter plot using `ggplot()` with `geom_smooth()`.  
- We also annotate the two teams `Oakland Athletics` and `Boston Red Sox`.  


---
\frametitle{Explore variable relationships: Least squared lines}

\tiny
```{r results="hold", fig.height = 3}
selected_teams <- c("Oakland Athletics", "Boston Red Sox")

data_agg %>%
  ggplot(aes(x = payroll_total, y = win_pct_ave)) + 
  geom_point(aes(color = team), size = 3) + 
  # only show names of selected_teams
  geom_text_repel(data = subset(data_agg, team %in% selected_teams),
                  aes(label = team, color = team), size = 5) +
  geom_smooth(method = "lm", formula = y ~ x, se = F,color = "red") + 
  geom_hline(aes(yintercept = mean(win_pct_ave)), color = "blue") +
  labs(title = "MLB Team's Overall Win  vs. Payroll", x = "Payroll", y = "Win") +
  theme_bw() +
  theme(legend.position = "none")
```


---
\frametitle{Conclusions/Discussions  }


*Answer to Question 1:*  

HERE is how the article concludes that Beane is worth as much as the GM in Red Sox. By looking at the above plot, Oakland A's win pct is more or less the same as that of Red Sox, so based on the LS equation, the team should have paid 2 billion!  

Do you agree with this argument? Why or why not?  

*Answer to Question 2:*  

From this regression line, we see a clear upward trend. Or precisely the least squared equation has a positive coefficient. Consequently, the more a team is paid the better performance we expect the team has.  


---
\frametitle{Conclusions/Discussions}


**Questions for you:**  

* Do you agree with the conclusions made based on a regression analysis shown above?  
* How would you carry out a study which may have done a better job? In what way?  

## Part II: Analyze pay and winning percent over time and by team

- Payroll and performance varies depending on teams and years.  
- We investigate changes over time and by teams to see how payroll relates to performance.  


---
\frametitle{Compare payroll and performance}


We can compare summary statistics of payrolls and performance among teams.

\tiny
```{r, payroll summary}
baseball %>% group_by(team) %>%
  summarise(payroll_mean = mean(payroll),
            win_pct_mean = mean(win_pct)) %>%
  arrange(-payroll_mean) %>%
  slice(1:5)
```
\normalsize
We see that *New York Yankees* has the highest payroll. *Boston Red Sox* is the next highest paid team. The effect of time is not included here. 


---
\frametitle{Compare payroll and performance}


Summary statistics can not describe the distributions of either payroll or performances. Back to back boxplots of payroll or winning percentage would capture the variability in details.

\tiny
```{r, back to back boxplots, fig.show='hide'}
baseball %>%
  ggplot(aes(x = team, y = win_pct, fill = team)) + 
  geom_boxplot() +
  xlab("Team") +
  ylab("Winning percentage") +
  ggtitle("Winning percentage by team") + 
  theme_bw() +
  theme(legend.position = "none",
        # adjust for margins around the plot; t: top; r: right; b: bottom; l: left
        plot.margin = margin(t = 5, r = 50, b = 5, l = 0, unit = "pt"), 
        axis.text.x = element_text(angle = -60, vjust = 0, hjust = 0))
```

---
\frametitle{Compare payroll and performance}


Summary statistics can not describe the distributions of either payroll or performances. Back to back boxplots of payroll or winning percentage would capture the variability in details.
```{r, back to back boxplots 2, fig.height = 3, echo=FALSE}
baseball %>%
  ggplot(aes(x = team, y = win_pct, fill = team)) + 
  geom_boxplot() +
  xlab("Team") +
  ylab("Winning percentage") +
  ggtitle("Winning percentage by team") + 
  theme_bw() +
  theme(legend.position = "none",
        # adjust for margins around the plot; t: top; r: right; b: bottom; l: left
        plot.margin = margin(t = 5, r = 50, b = 5, l = 0, unit = "pt"), 
        axis.text.x = element_text(angle = -60, vjust = 0, hjust = 0))
```

\normalsize
We see clearly that the medians/means and spreads are very different. Is there a more informative way to display this?


---
\frametitle{Compare payroll and performance  }


For example, we probably want to display the comparison by ranking the median:
\tiny
```{r set theme}
boxplot_theme <-
  theme_bw() +
  theme(legend.position = "none",
        # adjust for margins around the plot; t: top; r: right; b: bottom; l: left
        plot.margin = margin(t = 5, r = 50, b = 5, l = 0, unit = "pt"), 
        axis.text.x = element_text(angle = -60, vjust = 0, hjust = 0))
```

```{r, fig.show='hide'}
baseball %>%
  ggplot(aes(x = forcats::fct_reorder(team, -win_pct, .fun = median), #order win_pct in a decreasing order
             y = win_pct, fill = team)) + 
  geom_boxplot() +
  xlab("Team") +
  ylab("Winning percentage") +
  ggtitle("Winning percentage by team") + 
  boxplot_theme
```

---
\frametitle{Compare payroll and performance  }


We probably want to display the comparison by ranking the median for example:  
```{r, fig.height = 3, echo=FALSE}
baseball %>%
  ggplot(aes(x = forcats::fct_reorder(team, -win_pct, .fun = median), #order win_pct in a decreasing order
             y = win_pct, fill = team)) + 
  geom_boxplot() +
  xlab("Team") +
  ylab("Winning percentage") +
  ggtitle("Winning percentage by team") + 
  boxplot_theme
```

\normalsize
We see that `NY Yankees` and `Red Sox` are consistently good teams while `Oakland A's` has a good overall team performance but the performance varies.


---
\frametitle{Compare payroll and performance}

- Next: compare both `payroll` and `win_pct` by teams.  
- Let us try to line up two back to back boxplots together.  
- Notice that we tried to rank one variable while carrying the other variable in the same order.  
- The hope is to reveal the relationship between `payroll` and `performance`.  


---
\frametitle{Compare payroll and performance}

\tiny
```{r facet two b-b boxplots, fig.show='hide'}
# use reorder_within() and scale_x_reordered() from tidytext to order boxplot within each facet
library(tidytext)
p_win_pct <- baseball %>%
  ggplot(aes(x = forcats::fct_reorder(team, -win_pct, .fun = median), #order win_pct in a decreasing order
             y = win_pct, fill = team)) + 
  geom_boxplot() +
  xlab("Team") +
  ylab("Winning percentage") +
  ggtitle("Winning percentage by team") +
  boxplot_theme

p_payroll <- baseball %>%
  ggplot(aes(x = forcats::fct_reorder(team, -win_pct, .fun = median), #order win_pct in a decreasing order
             y = payroll, fill = team)) + 
  geom_boxplot() +
  xlab("Team") +
  ylab("Payroll") +
  ggtitle("Payroll by team in decreasing order of win_pct") + 
  boxplot_theme

gridExtra::grid.arrange(p_win_pct, p_payroll, ncol=1)
# ggpubr::ggarrange(p_win_pct, p_payroll, ncol = 1)
```

---
\frametitle{Compare payroll and performance}

```{r, fig.height = 4.5, echo=FALSE}
gridExtra::grid.arrange(p_win_pct, p_payroll, ncol=1)
```

Bingo! While `Oakland A's` payroll are consistently lower than that of `Red Sox`, they have similar performance!!!


---
\frametitle{Compare payroll and performance  }

Alternative boxplot faceting
\tiny
```{r, alternative facet b-b plots, eval=FALSE, fig.height = 6, fig.show='hide'}
# use reorder_within() and scale_x_reordered() from tidytext to order boxplot within each facet
library(tidytext)
# facet names
facet_names <- c("payroll" = "Payroll",
                 "win_pct" = "Winning percentage")
baseball %>%
  select(-win_num) %>%
  pivot_longer(cols = c("payroll", "win_pct"),
               names_to = "variable") %>%
  ggplot(aes(x = reorder_within(team, -value, variable, fun = median), 
             y = value, fill = team)) + 
  geom_boxplot() +
  scale_x_reordered() +
  facet_wrap(~ variable, ncol = 1, scales = "free",
             labeller = as_labeller(facet_names)) +
  xlab("Team") + ylab("") +
  ggtitle("Payroll and winning percentage by team") + 
  boxplot_theme
```


---
\frametitle{Compare payroll and performance  }

Alternative boxplot faceting
\tiny
```{r, alternative facet b-b plots 2, fig.height = 6, echo=FALSE}
# use reorder_within() and scale_x_reordered() from tidytext to order boxplot within each facet
library(tidytext)
# facet names
facet_names <- c("payroll" = "Payroll",
                 "win_pct" = "Winning percentage")
baseball %>%
  select(-win_num) %>%
  pivot_longer(cols = c("payroll", "win_pct"),
               names_to = "variable") %>%
  ggplot(aes(x = reorder_within(team, -value, variable, fun = median), 
             y = value, fill = team)) + 
  geom_boxplot() +
  scale_x_reordered() +
  facet_wrap(~ variable, ncol = 1, scales = "free",
             labeller = as_labeller(facet_names)) +
  xlab("Team") + ylab("") +
  ggtitle("Payroll and winning percentage by team") + 
  boxplot_theme
```


---
\frametitle{Comparing performance as a function of time  }

A time series of performance may reveal patterns of performance over the years to see if some teams are consistently better or worse. 
\tiny
```{r, fig.height = 2.5}
payroll_plot <- baseball %>%
  ggplot(aes(x = year, y = payroll, group = team, col = team)) + 
  geom_line() + 
  geom_point() +
  theme_bw() +
  ggtitle("Payroll versus year")
payroll_plot
```



---
\frametitle{Comparing performance as a function of time  }


A time series of performance may reveal patterns of performance over the years to see if some teams are consistently better or worse.
\tiny
```{r, fig.height = 2.5}
win_pct_plot <- baseball %>% 
  ggplot(aes(x = year, y = win_pct, group = team, col = team)) + 
  geom_line() + 
  geom_point() +
  theme_bw() +
  ggtitle("Winning percent versus year")
win_pct_plot
```


---
\frametitle{Comparing performance as a function of time  }


Winning pct plot with only `NY Yankees` (blue), `Boston Red Sox` (red)  and `Oakland Athletics ` (green) while keeping all other teams as background in gray. 
\tiny
```{r, fig.height = 3, fig.show='hide'}
selected_teams <- c("New York Yankees", "Boston Red Sox", "Oakland Athletics")

win_pct_plot <- baseball %>% 
  ggplot(aes(x = year, y = win_pct, group = team)) + 
  geom_line(col = "grey", alpha = .5) + 
  geom_point(col = "grey", alpha = .5) +
  geom_line(data = subset(baseball, team %in% selected_teams),
            aes(col = team)) +
  geom_point(data = subset(baseball, team %in% selected_teams),
            aes(col = team)) +  
  scale_color_manual(values = c("red", "darkblue", "darkgreen")) +
  theme_bw() +
  ggtitle("NY Yankees, Red Sox, Oakland A's")
  # layout(legend = list(x = 0.35, y = 0.99, orientation = 'h'))
win_pct_plot
```


---
\frametitle{Comparing performance as a function of time  }


Winning pct plot with only `NY Yankees` (blue), `Boston Red Sox` (red)  and `Oakland Athletics ` (green) while keeping all other teams as background in gray. 
```{r, fig.height = 3, echo=FALSE}
selected_teams <- c("New York Yankees", "Boston Red Sox", "Oakland Athletics")

win_pct_plot <- baseball %>% 
  ggplot(aes(x = year, y = win_pct, group = team)) + 
  geom_line(col = "grey", alpha = .5) + 
  geom_point(col = "grey", alpha = .5) +
  geom_line(data = subset(baseball, team %in% selected_teams),
            aes(col = team)) +
  geom_point(data = subset(baseball, team %in% selected_teams),
            aes(col = team)) +  
  scale_color_manual(values = c("red", "darkblue", "darkgreen")) +
  theme_bw() +
  ggtitle("NY Yankees, Red Sox, Oakland A's")
  # layout(legend = list(x = 0.35, y = 0.99, orientation = 'h'))
win_pct_plot
```

\normalsize
Now we see that `Red Sox` seems to perform better most of the time compared to the `Oakland A's`. 


---
\frametitle{Performance, Payroll and Year}

We are trying to reveal the relationship between `performance` and `payroll`. But it depends on which team at a given year. 
\tiny
```{r fig.height = 3, fig.show='hide'}
baseball %>%
  ggplot(aes(x=payroll, y=win_pct, group = team, color=team)) +
  geom_point()+
  geom_smooth(method="lm", formula=y~x, se=F,color = "red")+
  facet_wrap(~team) + 
  theme_bw() +
  theme(legend.position = "none") +
  ggtitle("`payroll` vs `win_pct` by team")
```


---
\frametitle{Performance, Payroll and Year}

```{r , echo=FALSE, fig.height = 4.5}
baseball %>%
  ggplot(aes(x=payroll, y=win_pct, group = team, color=team)) +
  geom_point()+
  geom_smooth(method="lm", formula=y~x, se=F,color = "red")+
  facet_wrap(~team) + 
  theme_bw() +
  theme(legend.position = "none") +
  ggtitle("`payroll` vs `win_pct` by team")
```

\normalsize
We see a discrepancy among teams for the relationship between `payroll` and `performance`. The positive trends vary from very positive to even negatively correlated.


---
\frametitle{Performance, Payroll and Year}

If we zoom in on a few teams we see a clear negative correlation between payroll and performance. What is missing here?
\tiny
```{r, fig.height = 3}
baseball %>% 
  filter(team %in% c("New York Yankees", "Boston Red Sox", "Oakland Athletics")) %>%
  ggplot(aes(x=payroll, y=win_pct, group = team, color=team)) +
  geom_point()+
  geom_smooth(method="lm", formula= y~x,  se=F,color = "red")+
  facet_wrap(~team) + 
  theme_bw() + 
  theme(legend.position = "bottom")
```


---
\frametitle{Performance, Payroll and Year}

We have seen before, `payroll` increases over years. It will be better to examine `payroll` v.s. `win_pct` by `year`:
\tiny
```{r, fig.height = 3, fig.show = 'hide'}
baseball %>%
  ggplot(aes(x=payroll, y=win_pct, group = year, color=team)) +
  geom_point()+
  geom_smooth(method="lm", formula=y~x, se=F,color = "red")+
  facet_wrap(~year) + 
  theme_bw() +
  theme(legend.position = 0)
```

---
\frametitle{Performance, Payroll and Year}

```{r, fig.height = 4, echo=FALSE}
baseball %>%
  ggplot(aes(x=payroll, y=win_pct, group = year, color=team)) +
  geom_point()+
  geom_smooth(method="lm", formula=y~x, se=F,color = "red")+
  facet_wrap(~year) + 
  theme_bw() +
  theme(legend.position = 0)
```

\normalsize
Now it seems to agree with our intuition, payroll and performance are indeed positively related for a given year. But the degree of relationship seems to change depending on which year and they are heavily controlled by some teams.


---
\frametitle{Performance, Payroll and Year}

We can summarize the above three dimension plots via a movie that tracks dynamic changes! 

\alert{See the plotly movie in the html file!}
\tiny
```{r, eval=FALSE}
selected_team <- c("Oakland Athletics", "New York Yankees", "Boston Red Sox")

p <- baseball %>%
  ggplot(aes(x=payroll, y=win_pct, color=team, frame = year)) +
   theme(legend.position = 0) + 
  geom_point() +
  geom_smooth(method="lm", formula=y~x, se=F, color = "red") + 
  geom_text(data = subset(baseball, team %in% selected_team),
            aes(label = team),
            show.legend = FALSE) +
  theme_bw()

ggplotly(p)
```
\normalsize
Perhaps we do not see strong evidence that `Oakland A's` is comparable to `Red Sox` in performance. 

# Conclusions and Discussion
- We have shown the power of exploratory data analysis to reveal correlation between payroll and performance.  
- Is payroll an important factor affecting the team performance if taking more factors into account? 
  - Here we assembled a dataset containing only performance and payroll at the team level over a span of 17 yrs.
- Analysis via aggregated statistics can be misleading. 
- Substantial variation can also exist within each team.
  - For example, payroll distribution is drastically different.
  - See this article on MLB income inequality:  
  [https://fivethirtyeight.com/features/good-mlb-teams-oppose-income-inequality/](https://fivethirtyeight.com/features/good-mlb-teams-oppose-income-inequality/)


---
\frametitle{Conclusions and Discussion}

Questions remain: 

1. Based on our current data, 
    a) what model will you consider to capture effects of payroll, year and team over the performace?
    b) would you use other measurements as dependent variable, e.g. annual payroll increase?

2. If you are asked to run the study to find out what are the main factors affecting performance, how would you do it?  
To narrow down the scope of the first step of the study, what information you may gather? 

# Appendix: Sample Statistics
We remind readers of the definition of sample statistics here.  

* Sample mean:

$$\bar{y} = \frac{1}{n}\sum_{i=1}^n y_i$$

* Sample variance:

$$s^2 = \frac{\sum_{i=1}^{n}(y_i - \bar{y})^2}{n-1}$$


---
\frametitle{Appendix: Sample Statistics}


* Sample Standard Deviation:

$$s = \sqrt\frac{\sum_{i=1}^{n}(y_i - \bar{y})^2} {n - 1}$$

* Sample correlation

$$r = \frac{\sum_{i=1}^n (x_i - \bar{x})(y_i - \bar{y})}{\sqrt{\sum_{i=1}^n (x_i - \bar{x})^2 \sum_{i=1}^n (y_i - \bar{y})^2}} = \frac{\sum_{i=1}^n (x_i - \bar{x})(y_i - \bar{y})}{s_x s_y} $$
