---
title: "WDS, HW 3"
author:
- Aaron
- Neil
- Ben
- Luke
- Tanvi
date: 'Due: 10:00PM,  July 21, 2021'
output:
  html_document:
    code_folding: show
    highlight: haddock
    theme: lumen
    toc: yes
    toc_depth: 4
    toc_float: yes
  pdf_document:
    toc: yes
    toc_depth: '4'
urlcolor: blue
---



```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, results = "hide", fig.width=8, fig.height=4)
options(scipen = 0, digits = 3)  # controls base R output
# check if you have ISLR package, if not, install it
if(!require('pacman')) {install.packages('pacman')}
pacman::p_load(ISLR, readxl, tidyverse, magrittr, dplyr, ggplot2, leaps, car, contrast) 
# ISLR will be installed and the that AUTO needed for this hw will be there
```


\pagebreak



# Case study: Automobiles efficiency

Are cars being built more efficient? Are Asian cars more efficient than cars built in America or Europe? To answer the questions we will use the `Auto` dataset from ISLR. The original dataset contains 408 observations about cars. It is similar to the CARS dataset that we use in our lectures. But it also collects information by years. To get the data, first install the package ISLR. The `Auto` dataset should be loaded automatically. The original data source is here: https://archive.ics.uci.edu/ml/datasets/auto+mpg

Get familiar with this dataset first. A good data set should be well documented. Use the command `?ISLR::Auto` to view a description of the dataset. Please add the variable list with names, brief descriptions and units of the variables below. 

```{r}
data1 <- Auto
str(data1)
names(data1)
summary(data1)
```

# EDA

Explore the data first. 

i. What is the range of `year`? Why is this important to know?
```{r}
year_range <- range(data1$year)
year_range
```
```{r}
data1 %>%
  select(year) %>%
  #range(data1$year, na.rm = TRUE)
  max(data1$year, na.rm=TRUE) - min(data1$year, na.rm=TRUE)
```

ii. Should `origin` be a continuous variable? Why or why not. In any case make `origin` a categorical variable.
```{r}
unique(data1$origin)

# factor origin
data1$origin <- factor(data1$origin, levels=3:1)
```
iii. Do you see any peculiarity in the data? 


## What effect does `time` have on `MPG`?

i. Show a scatter plot of `mpg` vs. `year` with the LS line imposed. Does the plot show a positive trend? 
```{r}
ggplot(data1, aes(x = year, y = mpg)) +
  geom_point(aes(color = origin)) +
  geom_smooth(method = "lm", se = FALSE) +
  geom_abline(intercept = 0, slope = 1) +
  theme_bw() +
  theme(legend.position = "none") +
  scale_x_continuous(breaks = seq(70, 90, 5)) +
  scale_y_continuous(breaks = seq(0, 200, 20)) +
  labs(x = "Years After 1900", y = "MPG")
```

ii. Now run a simple regression of `mpg` vs. `year` and report R's `summary` output. Is `year` a significant variable at the .05 level? State what effect `year` has on `mpg`, if any, according to this model. 
```{r}
lm1 <- lm(mpg ~ year, data=data1)
summary(lm1)
```

iii. Add `horsepower` on top of the variable `year` to your linear model. Is `year` still a significant variable at the .05 level? Give a precise interpretation of the `year`'s effect found here. 
```{r}
# add horsepower to the model
lm2 <- lm(mpg ~ year + horsepower, data = data1)
summary(lm2)
```

iv. The two 95% CI's for the coefficient of year differ among ii. and iii. How would you explain the difference to a non-statistician?

v. Create a model with interaction by fitting `lm(mpg ~ year * horsepower)`. Is the interaction effect significant at .05 level? Explain the year effect (if any). 
```{r}
lm3 <- lm(mpg ~ year* horsepower, data = data1)
summary(lm3)
```

<!-- ## Categorical predictors -->

<!-- Remember that the same variable can play different roles! Take a quick look at the variable `cylinders`, and try to use this variable in the following analyses wisely. We all agree that a larger number of cylinders will lower mpg. However, we can interpret `cylinders` as either a continuous (numeric) variable or a categorical variable. -->

<!-- a) Fit a model that treats `cylinders` as a continuous/numeric variable. Is `cylinders` significant at the 0.01 level? What effect does `cylinders` play in this model? -->

<!-- b) Fit a model that treats `cylinders` as a categorical/factor. Is `cylinders` significant at the .01 level? What is the effect of `cylinders` in this model? Describe the `cylinders` effect over `mpg`.  -->

<!-- c) What are the fundamental differences between treating `cylinders` as a continuous and categorical variable in your models?  -->


## Bring origin into the model

i. Do mpg's differ on average among different origin? Fit a linear model with mpg vs. origin. Report the output. 
```{r}
data1$origin <- factor(data1$origin, levels = c(3,2,1))
lm3 <- lm(mpg ~ origin, data = data1)
summary(lm3)
```

a) Are mpg's on average different among three regions? Perform a test at .01 level. When you reject the null hypothesis, what have you proved? 
```{r}
data1$origin <- factor(data1$origin, levels = c(3,2,1))
lm3 <- lm(mpg ~ origin, data = data1)
summary(lm3)

data1$origin <- factor(data1$origin, levels = c(1,2,3))
lm3 <- lm(mpg ~ origin, data = data1)
summary(lm3)

data1$origin <- factor(data1$origin, levels = c(2,1,3))
lm3 <- lm(mpg ~ origin, data = data1)
summary(lm3)
```

b) Describe on average which origin has the highest mpg and what it is. Which origin has the smallest mpg on average and what is it?
On an average origin 3 has the highest mpg and it is 30.451. The lowest is origin 1 and it is 20.033

c) Are Asian cars more efficient than American cars? Produce a 95% CI's for the difference. 


ii. Try to build a final model which includes `year` and `origin`. 

a) Describe the final model. Include diagnostic plots with particular focus on the model residuals and diagnoses.

b) Describe the year effect and origin effect in this final model.
Are cars being built more efficiently over time? Are Asian cars more efficient than cars built in America or Europe?

## Prediction

Use the final model to predict the `mpg` of the following car: A red car built in the US in 1983 that is 180 inches long, has eight cylinders, displaces 350 cu. inches, weighs 4000 pounds, and has a horsepower of 260. Also give a 95% CI for your prediction.








